---
title: "Annotation of MS-based Metabolomics Data"
package: MetaboCoreUtils
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Annotation of MS-based Metabolomics Data}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Metabolomics, Infrastructure, Quantitative }
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{Spectra,BiocStyle,msdata,MetaboAnnotation}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

**Package**: `r BiocStyle::Biocpkg("MetaboAnnotation")`<br />
**Authors**: `r packageDescription("MetaboAnnotation")[["Author"]] `<br />
**Compiled**: `r date()`


# Introduction

The `MetaboAnnotation` package defines high-level user functionality to support
and facilitate annotation of MS-based metabolomics data.


# Installation

The package can be installed with the `BiocManager` package. To
install `BiocManager` use `install.packages("BiocManager")` and, after that,
`BiocManager::install("MetaboAnnotation")` to install this package.


# Example use cases

The following sections illustrate example use cases of the functionality
provided by the `MetaboAnnotation` package.

```{r, message = FALSE}
library(MetaboAnnotation)
```


## Matching of m/z values

In this section a simple matching of feature m/z values against theoretical m/z
values is performed. This is the lowest level of confidence in metabolite anno-
tation. However, it gives ideas about potential metabolites that can be analyzed
in further downstream experiments and analyses.

The following example loads the feature table from a lipidomics experiments and
matches the measured m/z values against reference masses from LipidMaps. Below
we use a `data.frame` as *reference* database, but a
[`CompDb`](https://github.com/EuracBiomedicalResearch/CompoundDb) compound
database instance would also be supported.

```{r, message = FALSE}
ms1_features <- read.table(system.file("extdata", "MS1_example.txt",
                                       package = "MetaboAnnotation"),
                           header = TRUE, sep = "\t")
head(ms1_features)
target_df <- read.table(system.file("extdata", "LipidMaps_CompDB.txt",
                                    package = "MetaboAnnotation"),
                        header = TRUE, sep = "\t")
head(target_df)

## annotations <- matchMz(ms1_features, compDb,
##                        adducts = c("[M+H]+", "[M+Na]+"),
##                        tolerance = 0.005, ppm = 0)

```

For reference (target) compounds we have only the mass available. We need to
convert this mass to m/z values in order to match the m/z values from the
features (i.e. the query m/z values) against them. For this we need to define
the *most likely* ions/adducts that would be generated from the compounds based
on the ionization used in the experiment. We assume the most abundant adducts
from the compounds being `"[M+H]+"` and `"[M+Na]+`. We next perform the matching
with the `matchMz` function providing the query and target data as well as a
parameter object (in our case a `Mass2MzParam`) with the settings for the
matching. With the `Mass2MzParam`, the mass or target compounds get first
converted to m/z values, based on the defined adducts, and these are then
matched against the query m/z values (i.e. the m/z values for the features). To
get a full list of supported adducts the `MetaboCoreUtils::adductNames(polarity
= "positive")` or `MetaboCoreUtils::adductNames(polarity = "negative")` can be
used). Note also, to keep the runtime of this vignette short, we match only the
first 100 features.

```{r}
parm <- Mass2MzParam(adducts = c("[M+H]+", "[M+Na]+"),
                           tolerance = 0.005, ppm = 0)

matched_features <- matchMz(ms1_features[1:100, ], target_df, parm)
matched_features
```

From the tested 100 features 55 were matched against at least one target
compound (all matches are against a single compound). The result object (of type
`Matched`) contains the full query data frame and target data frames as well as
the matching information. We can access the original query data with `query` and
the original target data with `target` function:

```{r}
head(query(matched_features))
head(target(matched_features))
```

Functions `whichQuery` and `whichTarget` can be used to identify the rows in the
query and target data that could be matched:

```{r}
whichQuery(matched_features)
whichTarget(matched_features)
```

We can extract the full matching table with `data`. This returns a `DataFrame`
with all rows in *query* the corresponding matches in *target* along with the
matching adduct (column `"adduct"`) and the difference in m/z (column
`"score"`). Note that if a row in *query* matches multiple elements in *target*,
this row will be duplicated in the `DataFrame` returned by `data`. For rows that
can not be matched `NA` values are reported.

```{r}
data(matched_features)
```

See also the help page for `?Matched` for more details and information.


## Matching of MS/MS spectra

In this section we match experimental MS/MS spectra against reference
spectra. This can also be performed with functions from the
`r BiocStyle::Biocpkg("Spectra")` package (see
[SpectraTutorials](https://jorainer.github.io/SpectraTutorials/), but the
functions and concepts used here are more suitable to the *end user* as they
simplify the handling of the spectra matching results.

Below we load spectra from a file from a reversed-phase (DDA) LC-MS/MS run of
the Agilent Pesticide mix. With `filterMsLevel` we subset the data set to only
MS2 spectra. To reduce processing time of the example we further subset the
`Spectra` to a small set of selected MS2 spectra. In addition we assign *feature
identifiers* to each spectrum (again, for this example these are arbitrary IDs,
but in a *real* data analysis such identifiers could indicate to which LC-MS
feature these spectra belong).

```{r, message = FALSE}
library(Spectra)
library(msdata)
fl <- system.file("TripleTOF-SWATH", "PestMix1_DDA.mzML", package = "msdata")
pest_ms2 <- filterMsLevel(Spectra(fl), 2L)
## subset to selected spectra.
pest_ms2 <- pest_ms2[c(808, 809, 945:955)]
## assign arbitrary *feature IDs* to each spectrum.
pest_ms2$feature_id <- c("FT001", "FT001", "FT002", "FT003", "FT003", "FT003",
                         "FT004", "FT004", "FT004", "FT005", "FT005", "FT006",
                         "FT006")
## assign also *spectra IDs* to each
pest_ms2$spectrum_id <- paste0("sp_", seq_along(pest_ms2))
pest_ms2
```

This `Spectra` should now represent MS2 spectra associated
with LC-MS features from an untargeted LC-MS/MS experiment that we would like to
annotate by matching them against a spectral reference library.

We thus load below a `Spectra` object that represents MS2 data from a very small
subset of [MassBank](https://massbank.eu/MassBank/) release *2021.03*. This
small `Spectra` object is provided within this package but it would be possible
to use any other `Spectra` object instead (see the
[SpectraTutorials](https://jorainer.github.io/SpectraTutorials/) for different
ways and options to provide access to spectral libraries/databases *via*
`Spectra`).

```{r}
load(system.file("extdata", "minimb.RData", package = "MetaboAnnotation"))
minimb
```

We can now use the `matchSpectra` function to match each of our experimental
*query* spectra against the *target* (reference) spectra. Settings for this
matching can be defined with a dedicated *param* object. We use below the
`CompareSpectraParam` that uses the `compareSpectra` function from the `Spectra`
package to calculate similarities between each query spectrum and all target
spectra. `CompareSpectraParam` allows to set all individual settings for the
`compareSpectra` call with parameters `MAPFUN`, `ppm`, `tolerance` and `FUN`
(see the help on `compareSpectra` in the `r Biocpkg("Spectra")` package for more
details). In addition, we can *pre-filter* the target spectra for each
individual query spectrum to speed-up the calculations. By setting
`requirePrecursor = TRUE` we compare below each query spectrum only to target
spectra with matching precursor m/z (accepting a deviation defined by parameters
`ppm` and `tolerance`). By default, `matchSpectra` with `CompareSpectraParam`
considers spectra with a similarity score higher than 0.7 as *matching* and
these are thus reported.

```{r}
mtches <- matchSpectra(pest_ms2, minimb,
                       param = CompareSpectraParam(requirePrecursor = TRUE,
                                                   ppm = 10))
mtches
```

The results are reported as a `MatchedSpectra` object which represents the
matching results for all query spectra. This type of object contains all query
spectra, all target spectra, the matching information and the parameter object
with the settings of the matching. The object can be subsetted to e.g. matching
results for a specific query spectrum:

```{r}
mtches[1]
```

In this case, for the first query spectrum, no match was found among the target
spectra. Below we subset the `MatchedSpectra` to results for the second query
spectrum:

```{r}
mtches[2]
```

The second query spectrum could be matched to 4 target spectra. The matching
between query and target spectra can be n:m, i.e. each query spectrum can match
no or multiple target spectra and each target spectrum can be matched to none,
one or multiple query spectra.

Data (spectra variables of either the query and/or the target spectra) can be
extracted from the result object with the `spectraData` function or with `$`
(similar to a `Spectra` object). The `spectraVariables` function can be used to
list all available spectra variables in the result object:

```{r}
spectraVariables(mtches)
```

This lists the spectra variables from both the *query* **and** the *target*
spectra, with the prefix `"target_"` being used for spectra variable names of
the target spectra. Spectra variable `"score"` contains the similarity score.

We could thus use `$target_compound_name` to extract the compound name of the
matching target spectra for the second query spectrum:

```{r}
mtches[2]$target_compound_name
```

The same information can also be extracted on the *full* `MatchedSpectra`.
Below we use `$spectrum_id` to extract the query spectra identifiers we added
above from the full result object.

```{r}
mtches$spectrum_id
```

Because of the n:m mapping between query and target spectra, the number of
values returned by `$` (or `spectraData`) can be larger than the total number of
query spectra. Also in the example above, some of the spectra IDs are present
more than once in the result returned by `$spectrum_id`. The respective spectra
could be matched to more than one target spectrum (based on our settings) and
hence their IDs are reported multiple times. Both `spectraData` and `$` for
`MatchedSpectra` use a *left join* strategy to report/return values: a value
(row) is reported for each query spectrum (even if it does **not** match any
target spectrum) with eventually duplicated values (rows) if the query spectrum
matches more than one target spectrum (each value for a query spectrum is
repeated as many times as it matches target spectra). To illustrate this we
use below the `spectraData` function to extract specific data from our
result object, i.e. the spectrum and feature IDs for the query spectra we
defined above, the MS2 spectra similarity score, and the target spectra's ID and
compound name.

```{r}
mtches_df <- spectraData(mtches, columns = c("spectrum_id", "feature_id",
                                             "score", "target_spectrum_id",
                                             "target_compound_name"))
as.data.frame(mtches_df)
```

Using the `plotSpectraMirror` function we can visualize the matching results for
one query spectrum. Below we call this function to show all matches for the
second spectrum.

```{r}
plotSpectraMirror(mtches[2])
```

Not unexpectedly, the peak intensities of query and target spectra are on
different scales. While this was no problem for the similarity calculation (the
normalized dot-product which is used by default is independent of the absolute
peak values) it is not ideal for visualization. Thus, we apply below a
*normalization* function to both the query and target spectra and plot the
spectra again afterwards (see the help for `addProcessing` in the `Spectra`
package for more details on spectra data manipulations). This function will
replace the absolute spectra intensities with intensities relative to the
maximum intensity of the spectrum.

```{r}
norm_int <- function(x) {
    x[, "intensity"] <- x[, "intensity"] / max(x[, "intensity"], na.rm = TRUE)
    x
}
mtches <- addProcessing(mtches, norm_int)
plotSpectraMirror(mtches[2])
```

The query spectrum seems to nicely match the identified target spectra. Below we
extract the compound name of the target spectra for this second query spectrum.

```{r}
mtches[2]$target_compound_name
```

As alternative to the `CompareSpectraParam` we could also use the
`MatchForwardReverseParam` with `matchSpectra`. This has the same settings and
performs the same spectra similarity search than `CompareSpectraParam`, but
reports in addition (similar to MS-DIAL) to the (*forward*) similarity score
also the *reverse* spectra similarity score as well as the *presence ratio* for
matching spectra. While the default *forward* score is calculated considering
all peaks from the query and the target spectrum (the peak mapping is performed
using an *outer join* strategy), the *reverse score* is calculated only on peaks
that are present in the target spectrum and the matching peaks from the query
spectrum (the peak mapping is performed using a *right join* strategy). The
*presence ratio* is the ratio between the number of mapped peaks between the
query and the target spectrum and the total number of peaks in the target
spectrum. These values are available as spectra variables `"reverse_score"` and
`"presence_ratio"` in the result object). Below we perform the same spectra
matching as above, but using the `MatchForwardReverseParam`.

```{r}
mp <- MatchForwardReverseParam(requirePrecursor = TRUE, ppm = 10)
mtches <- matchSpectra(pest_ms2, minimb, param = mp)
mtches
```

Below we extract the query and target spectra IDs, the compound name and all
scores.

```{r}
as.data.frame(
    spectraData(mtches, c("spectrum_id", "target_spectrum_id",
                          "target_compound_name", "score", "reverse_score",
                          "presence_ratio")))
```

In these examples we matched query spectra only to target spectra if their
precursor m/z is ~ equal and reported only matches with a similarity higher than
0.7. `CompareSpectraParam`, through its parameter `THRESHFUN` would however also
allow other types of analyses. We could for example also report the *best
matching* target spectrum for each query spectrum, independently of whether the
similarity score is higher than a certain threshold. Below we perform such an
analysis defining a `THRESHFUN` that selects always the best match.

```{r}
select_top_match <- function(x) {
    which.max(x)
}
csp <- CompareSpectraParam(ppm = 10, requirePrecursor = FALSE,
                           THRESHFUN = select_top_match)
mtches <- matchSpectra(pest_ms2, minimb, param = csp)
res <- spectraData(mtches, columns = c("spectrum_id", "target_spectrum_id",
                                       "target_compound_name", "score"))
as.data.frame(res)
```


Note that this whole example would work on any `Spectra` object with MS2
spectra. Such objects could also be extracted from an `xcms`-based LC-MS/MS data
analysis with the `chromPeaksSpectra` or `featureSpectra` functions from the
`r Biocpkg("xcms")` package. Also,



# Session information {-}

```{r sessioninfo, echo=FALSE}
sessionInfo()
```
